---
title: "Introduction to GIS with R"
output: 
  md_document:
    variant: markdown
always_allow_html: yes
---

```{r setup, include=FALSE}
post <- "gis-with-r-intro"
source("common.r")
knitr::opts_chunk$set(fig.width = 8, fig.asp = 0.8)
```

The geographic visualization of data makes up one of the major branches of the Digital Humanities toolkit. There are a plethora of tools that can visualize geographic information from full-scale GIS applications such as [ArcGIS](https://www.esri.com) and [QGIS](https://qgis.org) to web-based tools like [Google maps](http://maps.google.com) to any number of programing languages. There are advantages and disadvantages to these different types of tools. Using a command-line interface [has a steep learning curve](https://jessesadler.com/post/excel-vs-r/),  but it has the benefit of enabling approaches to analysis and visualization that are customizable, transparent, and reproducible.[^1] My own interest in coding and [R](https://www.r-project.org) began with my desire to dip my toes into [geographic information systems (GIS)](https://en.wikipedia.org/wiki/Geographic_information_system) and create maps of an early modern correspondence network.  The goal of this post is to introduce the basic landscape of working with spatial data in R from the perspective of a non-specialist. Since the early 2000s, an active community of R developers has built a wide variety of packages to enable R to interface with geographic data. The extent of the geographic capabilities of R is readily apparent from the many packages listed in the [CRAN task view for spatial data](https://cran.r-project.org/web/views/Spatial.html).[^2]

In my previous post on [geocoding with R](https://jessesadler.com/post/geocoding-with-r/) I showed the use of the [`ggmap`](https://cran.r-project.org/web/packages/ggmap/index.html) package to geocode data and create maps using the [`ggplot2`](http://ggplot2.tidyverse.org) system. This post will build off of the location data obtained there to introduce the two main R packages that have standardized the use of spatial data in R. The[`sp`](https://cran.r-project.org/web/packages/sp/index.html) and [`sf`](https://cran.r-project.org/web/packages/sf/index.html) packages use different methodologies for integrating spatial data into R. The `sp` package introduced a coherent set of classes and methods for handling spatial data in 2005.[^3] The package remains the backbone of many packages that provide GIS capabilities in R.  The `sf` package implements the [simple features](https://cran.r-project.org/web/packages/sf/vignettes/sf1.html) open standard for the representation of geographic vector data in R. The package first appeared on [CRAN](http://cran.r-project.org) at the end of 2016 and is under very active development. The `sf` package is [meant to supersede `sp`](https://www.r-consortium.org/blog/2017/01/03/simple-features-now-on-cran), implementing ways to store spatial data in R that [integrate with the tidyverse workflow](https://edzer.github.io/UseR2017/ "Edzer Pebesma, Spatial Data in R: New Directions") of the [packages developed by Hadley Wickham and others](http://tidyverse.org).

There are a number of good resources on working with spatial data in R. The best sources for information about the `sp` and `sf` packages that I have found are Roger Bivand, Edzer Pebesma, and Virgilio Gómez-Rubio, *Applied Spatial Data Analysis with R* (2013) and the working book [Robin Lovelace, Jakub Nowosad, Jannes Muenchow, *Geocomputation with R*](https://bookdown.org/robinlovelace/geocompr/), which concentrate on `sp` and `sf` respectively. The [vignettes](https://cran.r-project.org/web/packages/sf/) for `sf` are also very helpful. The perspective that I adopt in this post is slightly different from these resources. In addition to more explicitly comparing `sp` and `sf`, this post approaches the two packages from the starting point of working with geocoded data with longitude and latitude values that must be transformed into spatial data. It takes the point of view of someone getting into GIS and does not assume that you are working with data that is already in a spatial format. In other words, this post provides information that I wish I knew as I learned to work with spatial data in R.  Therefore, I begin the post with a general overview of spatial data and how `sp` and `sf` implement the representation of spatial data in R. The second half of the post uses an example of mapping the locations of letters sent to a Dutch merchant in 1585 to show how to create, work with, and plot `sp` and `sf` objects. I highlight the differences between the two packages and ultimately discuss some reasons why the R spatial community is moving towards the use of the `sf` package.

## Spatial data and coordinate reference systems {#crs}
The distinguishing feature of spatial data is that it represents actual locations on Earth. To represent the geographic placement of an object you need two pieces of information: the coordinates of the object and a system of reference for how the coordinates relate to a physical location on Earth. The non-spherical shape of the Earth, which bulges at the equator, complicates the creation and use of a coordinate reference system or CRS and plethora of complex models have been created in attempts to accurately represent the Earth's surface. A CRS consists of one such ellipsoid or geometric model of the shape of the Earth and a datum, which identifies the origin and orientation of the coordinate axes on the ellipsoid, as well as the units of measurement.[^4] There are two types of CRSs: geographic and projected. Geographic reference systems represent points on an globe, using units of degrees longitude and latitude that correspond to angles measured from the center of the Earth as calculated using the given ellipsoid. A projected reference system uses a geometric model to project a 3-dimensional ellipsoid onto a 2-dimensional plane. A projection is necessary to create any 2-dimensional map, but it [results in the distortion](https://www.vox.com/world/2016/12/2/13817712/map-projection-mercator-globe) of aspects of the Earth's surface such as [area, direction, distance, and shape](http://axismaps.github.io/thematic-cartography/articles/projections.html). Despite the necessary distortion, projected reference systems are useful for geographic analysis, because they use linear units of measurement such as meters instead of degrees.[^5]

Between the variety of ellipsoidal models, the vast array of global and local datums, and the various kinds of projections, there are innumerable CRSs to choose from. The `sp` and `sf` packages implement two different ways to identify a specific CRS and to transform objects from one CRS to another: the [Proj.4 library](http://proj4.org) and [EPSG codes](http://www.epsg.org). The Proj.4 library uses an identification system of `+parameter=value` to define a CRS. The library uses a [wide range of parameters](http://proj4.org/parameters.html#parameter-list) to specify a CRS, but the most important are `+proj` and `+datum` for the projection and datum to be used. The Proj.4 parameters provides the basis for how `sp` and `sf` identify CRSs, but CRSs can also be identified through EPSG codes.[^6] The EPSG library supplies codes for well-known CRSs, and thus provides an easier method to identify a subset of the CRSs available through the Proj.4 library. You can access a data frame of over 5,000 EPSG codes available in R through the [`rgdal`](https://cran.r-project.org/web/packages/rgdal/index.html) package with the command `rgdal::make_EPSG()`. If you plan to deal with spatial data on a regular basis it is probably worth being familiar with widely used EPSG code such as 4326, which is geographic reference system that uses units of longitude and latitude on the [ World Geodetic System 1984 (WGS84) ellipsoid](https://en.wikipedia.org/wiki/World_Geodetic_System#WGS84). Two useful resources for looking up and obtaining information about specific EPSG codes are [EPSG.io](https://epsg.io) and [SpatialReference.org](http://spatialreference.org).

Spatial data with a defined CRS can represent either vector or raster data. Vector data is based on points — which can be connected to form lines and polygons — that are located within a coordinate reference system.  Raster data, on the other hand, consists of values within a grid system. For example, a road map is a vector data and a map using satellite imagery is raster data made up of pixels on a grid. `sp` has capabilities to work with both vector and raster data, while  `sf` and the simple features standard on which it is based only deals with vector data. This post will only discuss vector data, which is more widely used in the Humanities and Social Sciences.

## Overview of `sp` and `sf` packages {overview-spsf}
The complexity that comes with the identification and translation between the thousands of different coordinate reference systems makes it impossible for standard R objects such as data frames to properly represent spatial data. In addition, the best way to represent coordinates — in one or two columns — and the representation of more complex objects such as lines and polygons create problems for how to store spatial data. These shortcomings can be seen with geocoded data that contains longitude and latitude values in a data frame like that I created in [geocoding with R](https://jessesadler.com/post/geocoding-with-r/). Most geocoded data created through [Google Maps or other geocoding sites ](https://gis.stackexchange.com/questions/48949/epsg-3857-or-4326-for-googlemaps-openstreetmap-and-leaflet) provides coordinates in longitude and latitude values using the WGS84 ellipsoid and thus contains a CRS equivalent to the Proj.4 argument `+proj=longlat +datum=WGS84` or EPSG 4326. However, this information about the CRS is implicit and not contained anywhere within the data frame itself.

Due to the inadequacies of normal data frame objects to represent the variety of features of spatial data, the `sp` and `sf` packages both define their own classes of objects to store spatial data. From a beginner's perspective, the main differences between the `sp` and `sf` packages derive from the manner by which `sp` and `sf` classes store information about CRS, distinguish between points, lines, and polygons, and how they connect this spatial data to non-spatial data stored in a data frame.

The `sp` package uses what are known as [S4](https://bookdown.org/rdpeng/RProgDA/object-oriented-programming.html#s4) classes in R to represent spatial data. S4 objects are made up of **slots** that store different types of well-defined data. The slots can be accessed with the `@` symbol in the form of `object@slot`. The foundational `sp` class is the `Spatial` class, which has ten subclasses differentiated by the slots they contain. To get a better idea of what this all looks like, we can use the `getClass()` function, but first we have to load the `sp` package.[^7]

```{r sp classes}
# Load the sp package
library(sp)

# Spatial classes
getClass("Spatial")
```

The output lists the different `Spatial` classes and shows that the basis for all `Spatial` objects is the `bbox` and `proj4string` slots. The `proj4string` provides the CRS for an object through a Proj.4 definition, while the `bbox` slot provides a matrix of the minimum and maximum coordinates for the object. The `Spatial` subclasses add slots to the foundational slots of `bbox` and `proj4string` for the type of geometric data — points, lines, polygons, and grid for raster — and whether the object contains attribute data in the form of a data frame. We can see this from the slots in the `SpatialPoints` and the `SpatialPointsDataFrame` classes.

```{r sp slots names}
# slots for SpatialPoints class
slotNames("SpatialPoints")

# slots for SpatialPointsDataFrame
slotNames("SpatialPointsDataFrame")
```

The `SpatialPoints` class contains a `coords` slot to store the point coordinates. The `SpatialPointsDataFrame` class adds a `data` slot that enables points to be associated with attribute data in a data frame. The non-spatial data of a `Spatial*DataFrame` object can be accessed with `@data`. However, these S4-style classes do not fit within the definition of [tidy data](http://r4ds.had.co.nz/tidy-data.html) and cannot be manipulated using the [`dplyr`](http://r4ds.had.co.nz/transform.html) methods I have discussed in previous posts. Instead, all manipulation of the `@data` slot is done with base R commands. `sp` objects also cannot be plotted with `ggplot2`, but need to use the base or trellis plotting systems, though it is possible to force them into a compatible form with the [`broom`](https://cran.r-project.org/web/packages/broom/index.html) package.

In contrast to the `Spatial` class, the `sf` class — yes, the package and class have the same name — is an extension of data frames. Essentially, `sf` objects can be treated as data frames that also contain spatial data, as opposed to spatial data that may or may not also contain data frames. This enables `sf` objects to fit within the [tidyverse](http://tidyverse.org) workflow, making it possible to [manipulate them with `dplyr` commands](http://strimas.com/r/tidy-sf/). The [tidy](https://edzer.github.io/UseR2017/) nature of `sf` objects also means that they can can be [plotted with `ggplot2`](http://ggplot2.tidyverse.org/reference/ggsf.html), though currently this capability is only possible with the development version of `ggplot2`.

`sf` objects consist of rows of features, hence the name simple features, which have both non-spatial and spatial forms of data. The spatial data of an `sf` object is contained in a special geometry column that is of class `sfc`. The geometry column contains the same basic types of spatial data as the slots in `Spatial` objects: the CRS, coordinates, and type of geometric object. The `sfc` class has seven subclasses to denote the types of geometric objects within the geometry column, which are derived from the simple features standard. The possible geometric objects are point, linestring, polygon, multipoint, multilinestring, multipolygon, and geometrycollection for any combination of the other types.[^8]

With this basic introduction to spatial data and the `sp` and `sf` packages out of the way, the remainder of the post will demonstrate the creation and visualization of `Spatial` and `sf` objects by creating maps of the letters sent to Daniel van der Meulen in 1585. For this example, I want to map the total number of letters sent from and received in each location [as I did previously with `ggamp`](https://www.jessesadler.com/post/geocoding-with-r/#mapping-data). This will involve working with spatial data in the form of points, lines, and polygons. The points will represent the sources and destinations of the letters Daniel received and will be plotted on two different kinds of background maps using the [`rnaturalearth`](http://cran.r-project.org/web/packages/rnaturalearth) package to access coastal (lines) and country (polygons) world maps from the [Natural Earth](http://www.naturalearthdata.com) open-source repository of maps. In the process I will show how to make base R maps with `sp` and `ggplot2` plots with  `sf` and development version of `ggplot2`.

## Preparing the data {#preparing-data}
The first thing that we need to do is get information about the number of letters sent from and received in each city and join this non-spatial data with the longitude and latitude information that was created in the [geocoding with R](https://jessesadler.com/post/geocoding-with-r/) post. This will provide the necessary components to create spatial points object in both `sp` and `sf`. You can find the data and the R script that goes along with this example on [GitHub](https://github.com/jessesadler/intro-to-r). Let's start by loading the necessary packages and the two files containing data on the letters and the location of the cities in the letters data.

```{r packages and data}
# Load the packages
library(tidyverse)
library(sp)
library(sf)
library(rnaturalearth)

# Load the data
letters <- read_csv("data/correspondence-data-1585.csv")
locations <- read_csv("data/locations.csv")
```

The `letters` data frame contains 114 rows of letters with columns representing the writer, source, destination, and date of each letter. This example will only make use of the source and destination of the letters. The `locations` data frame contains columns for the name, longitude, and latitude of the 13 cities found in the letters data. There are a couple of ways that the data can be prepared, but in this case I want the data for the source and destination of the letters to be in a single data frame in order to minimize the objects that we will be dealing with in the post. The goal is to create a data frame with a column for the name of the city, the number of letters, and whether this information refers to letters sent or letters received. This can be done by separately finding the number of letters sent to each location and the number of letters received in each location and then joining the two data frames together.

We can get the number of letters sent from and received in each location by grouping the data by source or destination and using either `count()` or `summarise(count = n())` to count the number of occurrences in the chosen grouping. This provides us with a count column, but we also want to structure the data frames for source and destination in the same way. Therefore, I rename the column containing the city information to “place.” Doing this removes information about whether the observation is a source or destination, and so I add in this data with `add_column()`, creating a "type" column. The last step in this pipeline is to remove the grouping from the data frames, as this is no longer useful information and could influence later manipulations.

```{r prepare data}
# Letters per source
sources <- letters %>% 
  group_by(source) %>% 
  count() %>% 
  rename(place = source) %>% 
  add_column(type = "source") %>% 
  ungroup()

# Letters per destination
destinations <- letters %>% 
  group_by(destination) %>% 
  count() %>% 
  rename(place = destination) %>% 
  add_column(type = "destination") %>% 
  ungroup()
```

At this point, you can use either `rbind()` or `full_join()` to bind the rows of `sources` and `destinations` together to create a `letters_data` data frame that contains the non-spatial data we will be using. In addition, we need to make a small but important change to the type of data in the "type" column to make it possible to differentiate between the sources and destinations in the base R plots created below. We can transform the "type" column to a factor using `mutate()` and `as_factor()` from the [`forcats`](http://forcats.tidyverse.org) package. This changes the way that R represents the data behind the scenes such that "source" and "destination" can be represented by numeric integers, which is necessary for working with colors in base R plots.[^9]

```{r letters_data}
# Bind the rows of the two data frames
# and change type column to factor
letters_data <- rbind(sources, destinations) %>% 
  mutate(type = as_factor(type))
```

Let's look at the result. We can see that the data is in the format that we are looking for and that the "type" column has been correctly transformed into factors.

```{r print letters_data}
# Print letters_data
letters_data
```

Having created the data for the number of letters sent from and received in each location, we now need to add the longitude and latitude of the cities. We can do this with a left join with the `locations` data frame, using the "place" column as the key to link the two data frames.

```{r geocode data}
# Join letters_data to locations
geo_data <- left_join(letters_data, locations, by = "place")

# Print data with longitude and latitude columns
geo_data
```

The result is a data frame with 14 rows and 5 columns that give the number of letters sent from and received in the different locations and the longitude and latitude of those places. We now have all of the information that we need to transform this data into a spatial object with both `sp` and `sf`. We will then be able to map the points on base maps provided by the `rnaturalearth` package.

## Spatial data with `sp` {#sp}
The creation of a `Spatial` object involves providing data for the slots expected for each subclass. Creating a `SpatialPoints` object requires data for the `coords` and `proj4string` slots.[^10] The information for the `coords` slot can be filled by a data frame of the longitude and latitude values from `geo_data`. This can be done by creating a new data frame that only contains the "lon" and "lat" columns of `geo_data`. As noted above, the CRS for our data is implicitly longitude and latitude values on the WGS84 ellipsoid, which we can identify through the `CRS()` function containing a proj.4 argument. We can either specify the specific projection and datum arguments with `+proj=longlat +datum=WGS84` or use the EPSG code with `+init=epsg:4326`.

```{r points_sp}
# Create data frame of only longitude and latitude values
coords <- select(geo_data, lon, lat)

# Create SpatialPoints object with coords and CRS
points_sp <- SpatialPoints(coords = coords,
                           proj4string = CRS("+proj=longlat +datum=WGS84"))
```

Printing out our newly created `SpatialPoints` object does not make for particularly exciting reading. We see the coordinates, which are now in matrix form, and the CRS of the object. We can get a better insight into the structure of the object with the `str()` command, even if the output is a bit off-putting at first. This shows the three slots for the `SpatialPoints` object and the values contained within the slots.

```{r print SpatialPoints}
# Print SpatialPoints object
points_sp

# Structure of SpatialPoints object
str(points_sp)
```

We have now successfully created a `Spatial` object. However, `points_sp` does not contain any of the non-spatial data that was created in preparing the data. We can attach this non-spatial data to the `Spatial` data by creating a `SpatialPointsDataFrame` object. This is done in the same manner as a `SpatialPoints` object but also includes filling the `data` slot with the `letters_data` data frame.[^11] Because the data in both the `coords` and `data` slots derived from the same source, the coordinates and non-spatial data will be correctly aligned by row number.

```{r points_spdf}
# Create SpatialPointsDataFrame object
points_spdf <- SpatialPointsDataFrame(coords = coords,
                                      data = letters_data,  
                                      proj4string = CRS("+proj=longlat +datum=WGS84"))
```

Printing out `points_spdf` shows a result not dramatically different from printing out the `geo_data` data frame. However, inspecting the structure of the object shows that it has a very different composition.

```{r print SpatialPointsDataFrame}
# Print out SpatialPointsDataFrame
points_spdf

# Structure of SpatialPointsDataFrame object
str(points_spdf)
```

The nature of a `SpatialPointsDataFrame` object means that they cannot be manipulated in the same way as a normal data frame. You can access the data frame as you would any of the other slots in a `Spatial` object with the `points_spdf@data` notation. However, if you want to make changes to the `points_spdf` through an aspect of the `data` slot, you cannot use `dplyr` commands. Instead, you need to use base R methods, which can be a bit more convoluted. For instance, if you want to get the observations of places in which Daniel either received or was sent a certain number of letters you would use `@data` to access the data frame, `[]` to subset, and `$` to access the column that you want to subset. You can see an example of this below, creating a `SpatialPointsDataFrame` that contains the three locations in which "n" is greater than 10 and simply printing out the result to demonstrate the workflow.

```{r sp slots}
# Access the data frame of a SpatialPointsDataFrame
points_spdf@data

# Example of subsetting `points_spdf` to return locations with "n" greater than 10
points_spdf[points_spdf@data$n > 10, ]
```

To make a map with the `SpatialPoints` objects of the correspondence of Daniel, we need to get a background map of western Europe to provide geographical context. There are a number of R packages that provide access to various maps which can be read in as `Spatial` data. Here, I will make use of the `rnaturalearth` package to get access to a map of the coastlines of the world and another for the countries in the world.[^12] This will create two different kinds of `Spatial` objects. Because both maps contain non-spatial attribute data, they will be of class `SpatialLinesDataFrame` and `SpatialpolygonsDataFrame`.

```{r rnatural earth sp maps}
# Get coastal and country world maps as Spatial objects
coast_sp <- ne_coastline(scale = "medium")
countries_sp <- ne_countries(scale = "medium")
```

Not only are `coast_sp` and `countries_sp` useful for mapping the points data, they also provide an opportunity to look at the structure of more complex `Spatial` objects. Printing out either object fills up the entire console with an overflow of information. Instead, we can get an idea of the structure of the two objects with the `str()` command. To ensure that we do not overloaded with information with this command, I will only show the first two levels, which shows information about the slots but not on the data within the slots.

```{r look at sp background maps}
# Structure of SpatialLinesDataFrame
str(coast_sp, max.level = 2)

#Structure of SpatialPolygonsDataFrame
str(countries_sp, max.level = 2)
```

The output from `str()` shows that `coast_sp` and `countries_sp` are mainly distinguished by the presence of a `lines` or a `polygons` slot. Both of the objects contain data frames, but they are very different. The data frame in `coast_sp` only contains two columns, but `countries_sp` has 63 columns. This is because the countries map from `rnaturalearth` contains a variety of population, economic, and labeling data. We do not need to deal with the `data` slots from either of these objects, but they do provide a good example of what these kind of `Spatial` objects look like.

The last thing that we should do before plotting the points on our background maps is to make sure that our objects all have the same CRS. If `points_spdf` and `coast_sp` have different CRSs, the points will not be correctly plotted. We can check this with either the `proj4string()` function or printing out the `proj4string` slot. Checking this shows that all of the `Spatial` objects are using geographic coordinates on the WSG84 ellipsoid, enabling us to move forward with plotting.

```{r check CRS}
# Check CRS of Spatial objects
coast_sp@proj4string
proj4string(countries_sp)
proj4string(points_spdf)
```

## Mapping with `sp` {#mapping-sp}
Just as `Spatial` objects cannot be manipulated with `dplyr` commands, they cannot be plotted using `ggplot2`. `Spatial` objects can be plotted using the base R plot system or with its own plotting method that uses the Trellis plotting system with the `spplot()` command. Here, I will stick to the base plotting methods, but if you want to look at `spplot()` methods, Edzer Pebesma has created a good [gallery of sp plots](https://edzer.github.io/sp/). For a more in depth discussion on both plotting methods see Chapter 3 of Bivand, Pebesma, and Gómez-Rubio, *Applied Spatial Data Analysis with R*. I will discuss creating a map with the base R plotting system in some detail, because the commands are not particularly intuitive, and I have found it more difficult to find good information about making base plots compared to the use of `ggplot2`.

The plotting method for base R uses incremental commands to add different layers, allowing you to plot multiple types of data and add annotations such as legends and titles. The process can be somewhat tedious, and if you decide to change an aspect of the plot you need to rerun the entire string of commands. The necessity of building up every level of the plot with explicit commands has advantages and disadvantages when compared to `ggplot2`, which automates certain aspects of plot creation. For instance, base plots do not automate the creation of legends in the same way that `ggplot2` does. This makes the creation of legends more difficult, but because you have to construct each piece of the legend, its production can be more transparent than with `ggplot2`.

Base plotting methods enable you to manipulate a [wide variety of parameters](http://colinfay.me/intro-to-r/graphical-procedures.html) about your plots. You can see the whole list with `?par`. For this post, I will concentrate on some of the more widely used parameters to plot the points data we have created above on two different background maps. With these maps I want to visualize three different types of data: the location of the points represented in `points_spdf@coords`, whether the point represents a source or destination of the letters from the "type" column, and a representation of the amount of letters sent or received in the location from the "n" column. The latter two types of data can be represented through color and the size of the points respectively. Let's see how this works in practice.

Even before beginning to plot our `Spatial` objects, I want to set up the plotting environment by modifying the default margins to a smaller size that will work for all of these plots. The margins of a plot are identified by the `mar` or `mai` arguments within `par()`. Here, I will use `mar`, which provides the margins in units of lines. To change the margins we need to pass a vector of values for the bottom, left, top, and right margins. In this case, I will use `par(mar = c(1, 1, 3, 1)) ` to give some minimal margins and some room for a title at the top of the plot.

The location of the points of a `SpatialPoints` object are plotted automatically with `plot(points_spdf)`. However, the default for plotting `SpatialPoints` is to represent the points by an addition sign. We can change the symbol to represent the points with `pch` or the plotting character parameter. There are 25 available characters, and here I will use `pch = 20`, which is a solid circle.

Modifying the color and size of these circles is a bit trickier. The color of the circles can be adjusted with the `col` parameter according to the information in `points_spdf$type`. We are able to map the colors of the points to the "type" variable, because we earlier changed the column to be factors instead of characters. If this had not been done, the `plot()` command would fail. The default colors for the first two levels of factors are black and red, which [are not particularly pleasant to look at](https://bookdown.org/rdpeng/exdata/plotting-and-color-in-r.html). However, we can create a new color palette with the `palette()` function. Here, I am using the named colors "darkorchid" and "darkorange", while also adding some transparency to the points with the `alpha()` function from `ggplot2` since some of the points overlap.[^13] You can get a list of all of the named colors in R with `colors()`. The palette has to be created before beginning to make the plot, and this will change the color palette for all further base plots. You can return to the default palette with `palette("default")`.

The `cex` parameter is used to modify the size of the points that will be plotted. We can set the size of the points to `points_spdf$n`, but first we need to make some modifications. We do not want the point sizes to be the actual value in the "n" column. A point with `cex = 95` — the maximum number of letters in the data — would create a point larger then the plot. Therefore, we need to adjust the minimum and maximum size of the points according to some formula. This could be done in a variety of ways, but here I choose the formula `sqrt(points_spdf$n)/2 + 0.25`, which reduces the minimum `cex` to 0.75 and the maximum to a bit over 5. The other additions that I make are to add a box around the plot with `box()` and a title with `title()` in separate commands. These two commands decorate the plot and provides and example of the way that layers can be added to a base plot. Let's see how this looks.

```{r points_spdf-plot}
# Create a new color palette to distinguish source and destination
palette(alpha(c("darkorchid", "darkorange"), 0.7))
# Set margins for bottom, left, top, and right of plot
par(mar = c(1, 1, 3, 1))

# Plot points
plot(points_spdf,
     pch = 20,
     col = points_spdf$type,
     cex = sqrt(points_spdf$n)/2 + 0.25)

# Add a box around the plot
box()
# Add a title
title(main = "Correspondence of Daniel van der Meulen, 1585")
```

Ok, we have points, but we do not have a map. We can add a background map with a separate call to the `plot()` function that contains `add = TRUE`. One nice aspect of using the base plotting methods with `Spatial` objects is that even though the coastline map we are using is a world map, adding it to our previous plot properly scales the map to the geographical extent of the data from the first plotting function. The only other change is to make the color of the coastlines black with the `col` parameter. This needs to be done, because we changed the color palette above. If we did not make this change, the coastlines would be drawn in dark orchid.

The final step in making this map is to create legends to show the meaning of the colors and size of the points. This will require making two legends, which can be done with two calls to the `legend()` function. It is a bit fiddly to make these legends and build up every aspect, but the basis for what needs to be done is already present in the `plot()` functions. The `legend()` function takes a position for where to place the legend and the data to use for it. It is then a matter of using parameters to create informative legends.

Creating a legend for the distinction between sources and destinations uses data from the levels of the factors in the type column, which can be represented by `levels(points_spdf$type)`. We can map the colors to the levels with a vector containing the two levels, and here I use the shortcut of `1:2` to represent the two colors in the palette we created above. The `pch` parameter provides a symbol to represent the colors. In this case, I decided to use a filled square. The `pt.cex` parameter is used to increase the size of the colored squares, which we need to use because the `cex` parameter within `legend()` determines the size of the entire legend.

The legend to relate the size of the points to the number of letters sent from or received in each location is a bit more difficult to create, as we do not yet have all the necessary components for the legend. A legend for the size of the points will possess three or four points of different sizes and the number of letters they represent. `ggplot2` does this creation automatically, but with base plots you have to choose the number and size of the points yourself. This means we need to make our own data for the legend, which I do here by creating a `pointsize` vector that will be used to show the size of points representing 1, 50, and 100 letters.[^14] Once this is done, the creation of the legend is fairly straightforward. The key is to remember to make the point sizes for our vector of numbers equal to the points on the map by applying the same formula and using the same `pch`. A title for the legend makes it clearer that the map represents letters.[^15]

```{r coast_sp-plot}
# Pointsize vector for legend
pointsize <- c(1, 50, 100)
par(mar = c(1, 1, 3, 1))

# Plot points
plot(points_spdf,
     pch = 20,
     col = points_spdf$type,
     cex = sqrt(points_spdf$n)/2 + 0.25)
# Plot coastlines background map
plot(coast_sp,
     col = "black",
     add = TRUE)
# Add a box around the plot
box()

# Legend for colors
legend("topright", legend = levels(points_spdf$type),
       pt.cex = 2,
       col = 1:2,
       pch = 15)

# legend for size of points
legend("right", legend = pointsize,
       pt.cex = (sqrt(pointsize)/2 + 0.25),
       col = "black",
       pch = 20,
       title = "Letters")

# Title for the map
title(main = "Correspondence of Daniel van der Meulen, 1585")
```

The resulting map is maybe a bit sparse, but it provides all the necessary information on the locations and magnitude of Daniel's correspondence in 1585. One change that we might want to make is to color in the land to distinguish more clearly between land and water and to add some color to the plot. This cannot be done with our `SpatialLines` of the coastlines, which does not have a geometric interior that could be filled. Instead we need to use a `SpatialPolygons` object such as `countries_sp`. To make this second map we can reuse many of the parameters from our first map, but we do need to make one significant change to the plotting workflow. Because we want to fill in the land area of the map, we need to plot the background map before plotting the points. Otherwise the background map will cover the points. However, this disturbs the automatic subsetting of the world map that occurred in the first plot. We can get around this by changing the bounding box of `countries_sp` (`countries_sp@bbox`) to match that of `points_spdf` with the `bbox()` function.

```{r bbox for countries_sp}
# Make bounding box for countries_sp match
# bounding box of points_spdf
countries_sp@bbox <- bbox(points_spdf)
```

Once the countries map is properly limited to the extent of the points, we can reuse most of the parameters from the first map. The only difference from the previous plot other than placing the background map first in the plotting order is the choice of colors for `countries_sp`. For a `SpatialPolygons` object the `col` parameter adjusts the fill color for the polygons, and the `border` parameter modifies the color of the lines of the polygons. Using a map with modern country boundaries is clearly anachronistic for plotting letters from 1585, but the boundaries do help to contextualize the locations. If you want to obscure the country borders completely, you can make the `col` and `border` parameters the same color. In this case, I will use the `gray()` function, which enables you to choose a gray level from 0 to 1 with 0 representing black and 1 for white.

```{r countries_sp-plot}
par(mar = c(1, 1, 3, 1))

# Plot countries map and color with grays
plot(countries_sp,
     col = gray(0.8),
     border = gray(0.7))
# Plot points
plot(points_spdf,
     pch = 20,
     col = points_spdf$type, 
     cex = sqrt(points_spdf$n)/2 + 0.25,
     add = TRUE)
# Add a box around the plot
box()

# Legend for colors
legend("topright",
       legend = levels(points_spdf$type),
       pt.cex = 2,
       col = 1:2,
       pch = 15)
# legend for size of points
legend("right",
       legend = pointsize,
       pt.cex = (sqrt(pointsize)/2 + 0.25),
       col = "black",
       pch = 20,
       title = "Letters")

# Title for the map
title(main = "Correspondence of Daniel van der Meulen, 1585")
```

The above maps along with the creation of `Spatial` objects and the use of maps from outside sources demonstrates the value of the `sp` package. The above examples only show the basics of working with different kinds of `Spatial` objects and does not take advantage of any of the spatial transformations or calculations that using `sp` enables. However, the R GIS community is increasingly moving towards the `sf` package. The `sf` package provides almost all of the [capabilities of `sp`](https://github.com/r-spatial/sf/wiki/migrating), but it uses objects that are easier to work with than the S4-style classes of `sp`. The next section will replicate the workflow of creating and mapping spatial points data using `sf` methods, which will serve to illuminate the differences between the two packages.

## Spatial data with `sf` {#sf}
As I noted above, an `sf` object is a data-frame-like object that contains a special geometry column that is of class `sfc`. The geometry column stores the spatial forms of data such as the CRS, coordinates, and type of geometric object. The creation of an `sf` object with a geometry column of class `sfc_POINT` is similar to creating a `SpatialPointsDataFrame`, but it can be done in a single step with a data frame that contains longitude and latitude columns. The `st_as_sf()` function uses a vector of the coordinate columns for the `coords` argument and either a EPSG code or a Proj.4 definition for the `crs`.

```{r points_sf}
# Create sf object with geo_data data frame and CRS
points_sf <- st_as_sf(geo_data, coords = c("lon", "lat"), crs = 4326)
```

The `class()` function shows that `points_sf` is an `sf` object that is an extension of a `data.frame`, and in this case is an extension of `tbl_df`, because `geo_data` is a [tibble](http://r4ds.had.co.nz/tibbles.html). Printing `points_sf` results in an output similar to that of a tibble or data frame with additional information about the spatial aspects of the data provided by the "geometry" column. The information about the geometry shows the type of geometrical object, dimension, bounding box, and the CRS in both EPSG and proj.4 format when possible.

```{r print points_sf}
# Getting the class of an sf object shows that it is based on tibble and data frame.
class(points_sf)

# Printing out an sf object is similar to tibble or data frame
points_sf
```

You can inspect the class of the geometry column itself by using `$` to subset the geometry column. Or you can directly access information about the spatial data in the "geometry" column with the `st_geometry()` function.

```{r geometry of points_sf}
# Class of the geometry or sfc column
class(points_sf$geometry)

# Retrieve the geometry of an sf object
# to see coordinates, type of feature, and CRS
st_geometry(points_sf)
```

Having already loaded the Natural Earth coastline and countries maps as `Spatial` objects, there are two possibilities for getting the same data as `sf` objects. The `sf` package provides a method for converting from `Spatial` classes to `sf` with the `st_as_sf()` function that was used above to create an `sf` object from a data frame. It is useful to know that you can easily convert an object from `Spatial` to `sf` and back, but you will obviously not always have `Spatial` objects already created. The `rnaturalearth` package enables you to create `sf` objects with the `returnclass = "sf"` argument. Here, I will use the latter method.[^16]

```{r rnatural earth maps sf}
# Get coastal and country world maps as sf objects
coast_sf <- ne_coastline(scale = "medium", returnclass = "sf")
countries_sf <- ne_countries(scale = "medium", returnclass = "sf")
```

The data-frame-like nature of `sf` objects makes working with spatial data more transparent and in-line with other data workflows compared to the `sp` package. For starters, it is more reasonable to print out `sf` objects to the console, though the size of `coast_sf` and `countries_sf` makes this less than ideal. However, a big advantage of `sf` over `sp` is that you can easily peruse the contents of an object with the `View()` command or clicking on the object in the Environment pane in RStudio. With the `View()` command an `sf` object looks like any other data frame instead of a list of lists like a `Spatial` object. This makes it much easier to get an overview of the data that you are working with. Let's print out the first six rows, or features in the `sf` vocabulary, of `coast_sf` with the `head()` function to give an idea of what the object looks like. The structure of the output is similar to that of `points_sf` with a data frame containing a geometry column and information about the geometry of the object. In this case, we can see that `coast_sf` is of class `MULTILINESTRING`.

```{r print coast_sf}
# Print first six rows of coast_sf
head(coast_sf)
```

Unlike `Spatial` objects that use slots to distinguish the classes, all `sf` objects can generally be treated as data frames when wrangling the data. The only difference in the structure of `points_sf` compared to `coast_sf` or `countries_sf` is the class of the geometry column. This makes it possible to use workflows from the tidyverse. For instance, returning the subset of locations from `points_sf` that have a "n" greater than 10, as I showed above with `points_spdf`, can be done with `dplyr`'s `filter()` function.

```{r filter sf}
# Subset of locations with "n" greater than 10
filter(points_sf, n > 10)
```

The advantages of using `dplyr` to manipulate spatial objects is more apparent when working with complex objects such as `countries_sf` and using the pipe (`%>%`) to link commands. For example, if you want to look at the countries in South America, you can much more easily navigate the 241 rows and 63 columns of non-spatial data in the `sf` version of the countries world map than the `sp` version. Printing out the names of the columns with `colnames()` shows that we can filter countries in South America through the "continent" column. We can simplify the non-spatial data by selecting only the "name" column to identify the countries. Even more impressively, it is possible to combine data wrangling commands with spatial transformations such as changing the CRS of the object with `st_transform()`. As a demonstration, I can change the CRS of the subset of `countries_sf` to use the [Mollweide projection](https://en.wikipedia.org/wiki/Mollweide_projection), which prioritizes accuracy of area over shape and angle. The result of the pipeline of commands is a legible `sf` object with 13 rows for the countries in South America with a new CRS. Note that because there is no EPSG code for the Mollweide projection the `epsg` of the new object is `NA`, but the `proj4string` is changed to `+proj=moll`. Notice too that the coordinates have been changed from degrees to meters.

```{r South American countries}
# South American countries with new CRS
countries_sf %>% 
  filter(continent == "South America") %>% 
  select(name) %>% 
  st_transform(crs = "+proj=moll +datum=WGS84")
```

We can even visualize this transformation of `countries_sf` by piping directly to a base `plot()` command. Like `sp`, `sf` has its own plotting methods, which differ slightly from the base plotting system.[^17] The changes to the defaults that I make here are to remove the automatically created legend of country names with `key.pos = NULL` and provide a title for the plot. I also add graticules, or grid lines along longitude and latitude, to more clearly show the effects of the change in projection. Before making the map, we need to return the color palette to its default settings to prevent the borders of the countries from being drawn in dark orchid.

```{r sa-countries, fig.asp = 1}
# Return to default palette
palette("default")

# Map of South American countries
countries_sf %>% 
  filter(continent == "South America") %>% 
  select(name) %>% 
  st_transform(crs = "+proj=moll +datum=WGS84") %>% 
  plot(key.pos = NULL, graticule = TRUE, main = "South America")
```

## Mapping with `sf` and `ggplot2` {#mapping-sf}
Returning to the example of the correspondence of Daniel van der Meulen, it would be possible to use `sf` to recreate the base R plots made with the `sp` package. However, another advantage of `sf` is that it works with `ggplot2`, and so I will concentrate on making maps using the `ggplot2` system similar to those I made with `ggmap` in [a previous post](https://www.jessesadler.com/post/geocoding-with-r/#mapping-data). `ggplot2` implements the plotting of `sf` objects through the creation of a specific geom, `geom_sf()`. `geom_sf()` is only available in versions of `ggplot2` that are greater than 2.2.1, which is the current CRAN version. Therefore, you need to download the development version of `ggplot2` with `devtools::install_github("tidyverse/ggplot2")` to plot `sf` objects.

The manner in which `ggplot2` creates a plot through a single command rather than a series of separate commands as in base plots means that we need to explicitly deal with the incongruence in the geographic extent of `points_sf` and the background maps. If you make a `ggplot2` plot with `points_sf` and `coast_sf` without modifications, you will get a wold map with some small points in western Europe. There are two ways to deal with this incongruence by either changing the bounding box of the background map before making the plot or within the plot itself. As far as I know, there is no equivalent method in the `sf` package to the command we used above to modify the `bbox` slot of the `countries_sp` object. However, it is possible to geographically subset an `sf` object by a bounding box of coordinates using the `ms_clip()` function from the [`rmapshaper`](https://cran.r-project.org/web/packages/rmapshaper/) package. The other option is to leave `coast_sf` and `countries_sf` as they are and adjust the geographic extent of the objects in the plot command itself with the `coord_sf()` function and the `xlim` and `ylim` arguments. Here, I will use the latter method.

There are some oddities in using `geom_sf()` compared to other geoms due to the character of `geom_sf()`. By its nature, `geom_sf()` is unlike other geoms in that it can be used to create points, lines, or polygons depending on the contents of the `sf` object. This variability results in the need to declare the geometrical object used in a `show.legend` argument for the legend to be properly formatted. Another difference that I have found from other geoms is that `geom_sf()` does not properly identify `sf` objects unless you explicitly identify them with `data = sf_object`. While these issues are something to be aware of, they do not substantively affect the the plots that we can create.

Let's start by creating a simple plot using `ggplot2` with `points_sf` and `coast_sf`. Note the explicit vocabulary to identify the `sf` objects as the data and the use of `show.legend = "point"` for the legend. Otherwise the plot recreates the basic structure of the above base R plots by mapping the aesthetics of `color` to the "type" variable and `size` to "n". Some transparency is added to all of the points with the `alpha` argument. The last change is to subset the geographic region of the plot, and thus of `coast_sf` to a chosen bounding box with `coord_sf()`. You can experiment with different values for `xlim` and `ylim` to see what works best.

```{r ggplot-demo}
ggplot() + 
  geom_sf(data = coast_sf) + 
  geom_sf(data = points_sf,
          aes(color = type, size = n),
          alpha = 0.7,
          show.legend = "point") +
  coord_sf(xlim = c(-1, 14), ylim = c(44, 55))
```

The map created by `ggplot2` shows the same information as the base plots. It includes legends by default so that we do not need to build them up from scratch, though with `geom_sf()` we do need to use `show.legend = "point"` to format the legends properly. The aesthetics of this map are quite different from the base plots and serves to show the defaults that come with the use of `geom_sf()` in `ggplot2`. The defaults for `geom_sf()` use the iconic `ggplot` gray background, but instead of the usual white grid lines, graticules for the longitude and latitude are shown. Notice that the tick marks for the plot are correctly labeled in units of degrees longitude and latitude and that the axes are not labeled, as the context makes clear that the x-axis is latitude and the y-axis is longitude — at least if you remember which one is which.

The basic structure of what we want is present in the above map, but we can improve upon it by adding some information and making aesthetic changes to the map. We can simplify the map and make it more similar to the above base R plots by removing the graticules, which can be done by passing `NA` to the datum argument within the `coord_sf()` function. On the other hand, it is possible to add information to the plot by labeling the points with `geom_text_repel()` from the [`ggrepel`](https://cran.r-project.org/web/packages/ggrepel/index.html) package, which will ensure that the labels for the points do not overlap. `geom_text_repel()` does not work with `sf` objects, but because we have made no geographic transformations to the data, we can use the data from the `locations` data frame to map the names of the locations to the longitude and latitude values.

Further changes to the non-data aspects of the map can be made through the [theme system used by `ggplot2`](http://r4ds.had.co.nz/graphics-for-communication.html). We can alter the look of the map by choosing a different built in theme, changing individual elements of the aesthetics, or a combination of the two. You can play around with the theming system to create the look you want. In this case, I will make changes to the theme and to individual aspects of the plot. Changing the theme to `theme_minimal()` changes the background to white, but it also adds axes labels, which we will want to remove. The labels for the axes, legends, and plot as a whole can be modified within the `labs()` function. To remove the labeling of the axes for the plot I set the `x` and `y` labels to `NULL`. I also change the labels for the legends to give a more descriptive name to the size legend and use title case for the color legend. The last change I make is to increase the size of the points within the color legend with the `guides()` function, increasing the size of the points to 6.

```{r ggplot-coastline}
# Load ggrepel package
library(ggrepel)

ggplot() + 
  geom_sf(data = coast_sf) + 
  geom_sf(data = points_sf, 
          aes(color = type, size = n), 
          alpha = 0.7, 
          show.legend = "point") +
  coord_sf(xlim = c(-1, 14), ylim = c(44, 55),
           datum = NA) + # removes graticules
  geom_text_repel(data = locations, 
                  aes(x = lon, y = lat, label = place)) +
  labs(title = "Correspondence of Daniel van der Meulen, 1585",
       size = "Letters",
       color = "Type",
       x = NULL,
       y = NULL) +
  guides(color = guide_legend(override.aes = list(size = 6))) +
  theme_minimal()
```

The result is a map that is much cleaner, and because of the points are labeled, more informative. We can reuse the thematic tweaks to create a plot with the countries background map to make it possible to fill in the land area as we did above. I will use the same gray scale for the colors to fill the polygons and for the lines of the polygons. The other change that I make in this plot is to use the `theme_bw()`, which adds a border around the plot.

```{r ggplot-countries}
ggplot() + 
  geom_sf(data = countries_sf,
          fill = gray(0.8), color = gray(0.7)) + 
  geom_sf(data = points_sf, 
          aes(color = type, size = n), 
          alpha = 0.7, 
          show.legend = "point") +
  coord_sf(xlim = c(-1, 14), ylim = c(44, 55),
           datum = NA) + # removes graticules
  geom_text_repel(data = locations, 
                  aes(x = lon, y = lat, label = place)) +
  labs(title = "Correspondence of Daniel van der Meulen, 1585",
       size = "Letters",
       color = "Type",
       x = NULL,
       y = NULL) +
  guides(color = guide_legend(override.aes = list(size = 6))) +
  theme_bw()
```

## Conclusion {#conclusion}
This post has only scraped the surface of the power of the `sp` and `sf` packages and the more general GIS capabilities of R. My goal with this post has been to provide an understanding of the nature of spatial data, how `sp` and `sf` implement the creation of spatial objects in R, and some of the consequences of the implementation details for modifying and plotting `Spatial` and `sf` objects. `sp` and its complimentary packages have long provided a foundation to make R a viable GIS platform. The recent development of the `sf` package has modernized the implementation of spatial data in R and made it possible to integrate spatial data into the tidyverse and `ggplot2` plotting system. `sf` has made it easier to work with spatial data in R by minimizing the distinction between spatial data and other forms of data you might deal with in R. There still remain uses for the `sp` package, and I think it is helpful to see the distinction between the two packages to better understand both, but `sf` is clearly the future for the representation of geographic vector data in R.

## Additional resources for GIS with R {#gis-resources}
- The CRAN task view for [Analysis of Spatial Data](https://cran.r-project.org/web/views/Spatial.html) provides a good overview of the variety of packages that deal with GIS.
- `sp`
	  - The best resource for learning more about the `sp` package is Roger Bivand, Edzer Pebesma, and Virgilio Gómez-Rubio, *Applied Spatial Data Analysis with R* (2013).
	  - There are a number of good tutorials on working with `Spatial` objects and a particularly useful resource is [Nick Eubank's, GIS in R](http://www.nickeubank.com/gis-in-r/).
- `sf`
	  - The [package vignettes for `sf`](https://cran.r-project.org/web/packages/sf/) are very helpful for providing an introduction to the package.
	  - The currently in development book [Geocomputation with R by Robin Lovelace, Jakub Nowosad, Jannes Muenchow](https://bookdown.org/robinlovelace/geocompr/) is already an invaluable resource and was of great use for writing this post.
	  - The [R Spatial blog](http://r-spatial.org) is a good place to keep up on developments of `sf` and complimentary packages that have adopted the use of `sf` objects.

[^1]:	[Edzer Pebesma, Daniel Nüst, and Roger Bivand, "The R Software Environment in Reproducible Geoscientific Research," *Eos* 93 (2012): 163–164.](http://onlinelibrary.wiley.com/doi/10.1029/2012EO160003/abstract)

[^2]:	For a good general introduction to the use and history of GIS with R, see the working book [Robin Lovelace, Jakub Nowosad, Jannes Muenchow, *Geocomputation with R*](https://bookdown.org/robinlovelace/geocompr/).

[^3]:	[E. J. Pebesma and R. S. Bivand, "Classes and methods for spatial data in R," *R News* 5, no. 2 (2005): 9–13.](https://cran.r-project.org/doc/Rnews/Rnews_2005-2.pdf)

[^4]:	For more information on CRSs see [Lovelace, Nowosad, and Muenchow, Geocomputation with R](https://geocompr.robinlovelace.net/spatial-class.html#crs-intro) and Bivand, Pebesma, and Gómez-Rubio, 84–91.

[^5]:	The distance between degrees of longitude and latitude are only equal at the equator. Lines of longitude get progressively closer together as they approach the poles where they all meet. This variance means you cannot calculate distance between points with longitude and latitude values without the use of complex geometric models.

[^6]:	The details of how `sp` and `sf` interface with the Proj.4 library are different, but this should not be something that most users will have to worry about. [ See the discussion in `sf` issue 280](https://github.com/r-spatial/sf/issues/280).

[^7]:	See chapter 2 of Bivand, Edzer Pebesma, and Virgilio Gómez-Rubio for a more in depth discussion of `Spatial` objects.

[^8]:	See the vignette [1. Simple Features for R](https://cran.r-project.org/web/packages/sf/vignettes/sf1.html) and the very good discussion of `sf` objects in chapter 2 of [Lovelace, Nowosad, and Muenchow, *Geocomputation with R*](https://geocompr.robinlovelace.net/spatial-class.html#sf-classes).

[^9]:	On factors see [chapter 15 of Garrett Grolemund and Hadley Wickham, *R for Data Science*](http://r4ds.had.co.nz/factors.html) and [Amelia McNamara and Nicholas J Horton, "Wrangling Categorical Data in R"](https://dx.doi.org/10.7287/peerj.preprints.3163v2).

[^10]:	The `bbox` slot will be created from the information provided by the other two slots.

[^11]:	I could use the `geo_data` data frame, but this would unnecessarily result in having the coordinates in both the `data` and `coords` slots.

[^12]:	You may need to download these maps the first time that you use them.

[^13]:	Instead of choosing your own colors, you could use any of the number of R packages that provide color palettes such as [`RColorBrewer`](https://cran.r-project.org/web/packages/RColorBrewer/index.html), but for this example I will keep it simple and make my own palette.

[^14]:	Make sure to create the `pointsize` vector before beginning the chain of `plot()` functions.

[^15]:	I rerun the `par()` command for the margins to ensure that the margins stay consistent across these plots.

[^16]:	The `sf` package has the ability to read in and convert a number of spatial data types used in GIS with `st_read()`. [See the vignette on reading and writing simple features](https://cran.r-project.org/web/packages/sf/vignettes/sf2.html).

[^17]:	The details of plotting `sf` objects are covered in a [package vignette](https://cran.r-project.org/web/packages/sf/vignettes/sf5.html).